---
title: "Supplementary-material"
author: "Fabio Favoretto, Joy Kumagai, Octavio Aburto, Alex Rogers"
date: "Last updated `r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
---


<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```



# Premise

This supplementary material is associated with figure X, XX, and XXX of main text. By following this supplementary it is possible to replicate in full the both figures and the analysis. 

In this supplementary, we assume some basic knowledge of the R programming language. If instructions are followed, this should be fully reproducible using R studio. For further comments on the results of this analysis please refer to the main paper, or please find [Contacts][Contacts] to get in touch with the authors to report comments, bugs or problems. 

Data and R code scripts to replicate figures and analysis can be downloaded from: 

A wrangled dataset of all the data mentioned in Table SI and SII is available in the `data` folder.

The data file `BP-10_dataset.xlsx` has three sheets: 

- `habitat_data`: which contains all the habitat extensions in squared kilometers for each country, as well as the area of each country Economic Exclusion Zone (EEZ) and total MPA area both in squared kilometers;
- `pressures`: which contains the environmental pressures, the total GDP (Gross Domestic Product in USD) and the estimated biodiversity for each of the country);
- `protected_area`: which contains how much of each habitat within each country is protected (protected);


# Data Sources

Beware that while all data used are open source (except for kelp which source has been removed until the dataset is published), specific permission to reuse and publish them are needed from data providers. Credit for the use of those data should also go to the proper source listed in the table SI and SII. 

## Habitat area 

We obtained the polygons of 6 habitats closely associated with the coast and 6 more closely associated with open ocean, for a total of 12 majour habitats (Table SI). We then calculated the area for each habitat by dissolving the resulting layer by Country and projecting it into the World Cylindrical Equal Area projection, and then using the "Calculate Geometry" tool in ArcGIS 10.5 Desktop software. Losses and gains in habitat extension that occurred the data aquisition dates were not accounted for. These data are available in the available BP-10_Dataset.xlsx file under the `habitat_data` sheet. 

*Table SI: habitat data type, date of creation, and detailed references*

| Habitat |  Date of Data | Data Type | Source |
| :---    |     :----:    |   :----:  |   ---: |
| Estuaries | 2003 | Polygon | Alder (2003) |
| Mangroves	| 1997 - 2000 | Polygon | Giri, et al. (2011) | 
| Saltmarsh	| 1973 - 2015 | Points | McOwen, et al. (2017) |
| Seagrasses | 1934 - 2015 | Polygon | UNEP-WCMC, Short FT (2017) |
| Coral Reefs | 1954 - 2018 |	Polygon | UNEP-WCMC, WorldFish Centre, WRI, TNC (2018) |
| Kelp | NA	| Point | Jorge Assis (submitted for publication) |
| Cold Corals | 1915 - 2014 |	Point	| Freiwald A (2017) |
| Sills | 1950-2009 | Polygon | Harris et al. (2014) |
| Seamounts/Guyots | 1950-2009 | Polygon | Harris et al. (2014) |
| Bridges | 1950-2009 | Polygon | Harris et al. (2014) |
| Rift Valleys | 1950-2009 | Polygon | Harris et al. (2014) |
| Hydrothermal Vents | 1994-2019 | Point | Beaulieu, S.E., Szafranski, K. (2019) |


The other variables used in this study and featured in the `pressures` sheet of the dataset are described in Table SII. 

*Table SII: other variables used in this study* 

| Variable      | Variable name | Description | Source     |
| :---      |    :----:   | :----: |         ---: |
| Exclusive Economic Zone | eez_area | Area of the EEZ for each country in squared kilometers | Sala *et al.,* 2018 |
| Marine Protected Area | mpa | MPA extension in squared kilometers in each country. The dataset was filtered by MPAs whose status was either designated, inscribed, adopted or established, thus removing not reported and proposed categories | We obtained MPA extension from the World Database of Protected Areas (UNEP-WCMC accessed in February 2019) |
| Species Biodiversity | biodiversity_points | Extracted value of estimated species diversity | raster from Reygondenau *et al., in press* |
| Pressures on the Marine Environment | pressures | The ecological and social factors that decrease health status | Ocean Health Index: http://www.oceanhealthindex.org/ | 
| Gross Domestic Product per capita | gdp_total | total GDP per country (current USD) | World Bank Open data: https://data.worldbank.org/ | 
| Protected habitat area | protected | Overlap between protected area and the target habitat in squared kilometers | This paper |


We calculated the protected habitat area by intersecting the dissolved MPA layer per each country EEZ for each of the habitat listed in Table SI. The resulting data are reported in the dataset under the sheet `protected_area`.
It needs to be clarified that being inside a MPA does not mean the habitat is protected, since the MPA objective and regulamentation might not involve the habitat at all. However, we consider that being inside an environmentally managed area should provide at least some indirect benefits to the habitat conservation.

---




Below we reproduce the R code used to produce figure X, XX, and XXX step by step. 

The R code (v.3.6.1) was written using R-studio IDE (v.1.2.1511), as well as this document, using the following packages that can be installed in R or through R-studio using the following commands: 

- `install.packages("tidyverse")`
- `install.packages("readxl")`
- `install.packages("cowplot")`
- `install.packages("ggthemes")`
- `install.packages("broom")`

Libraries needed:

```{r}

library(dplyr)
library(ggplot2)
library(readxl)
library(ggpubr)
library(broom)
library(ggthemes)

```


Loading data:

```{r}

pressures <- read_excel("data/BP-10-Dataset.xlsx", sheet = "pressures") 

```

First of we analysed some relationship with the pressures data used.


Human Pressure *vs* Species Biodiversity model

```{r}
tidy(lm(pressures_spp~biodiversity_points, data=pressures))
```


% MPA/EEZ *vs* % of World GDP model

```{r}
pressures2 <- pressures %>% 
        select(country, eez_area, mpa, gdp_tot) %>% 
        na.omit() %>% 
        mutate(percent_GDP = (gdp_tot/max(gdp_tot))*100) %>% 
        mutate(percent_MPA = (mpa/eez_area)*100) 

tidy(lm(percent_MPA ~ percent_GDP, data=pressures2))
```

Species Biodiversity *vs* % MPA/EEZ model

```{r}
pressures3 <- pressures %>% 
        select(country, eez_area, mpa, gdp_tot, biodiversity_points) %>% 
        na.omit() %>%
        mutate(percent_MPA = (mpa/eez_area)*100) 

tidy(lm(biodiversity_points ~ percent_MPA, data=pressures3))

```

## Figure X: 

To reproduce the scatterplots relating Biodiversity, protection and GDP one can use the following code: 

```{r}
pA <- pressures %>% 
      select(country, eez_area, mpa, pressures_spp, biodiversity_points) %>% 
      na.omit() %>% 
      mutate(pressures_spp = pressures_spp/max(pressures_spp)) %>% 
      ggplot(aes(x=biodiversity_points, y=pressures_spp)) +
      geom_point() +
      geom_smooth(method='lm',formula=y~x) +
      theme_light() +
      #annotate("text", x = 50, y = 1, label = "A", size=5 ,fontface="bold") +
      labs(x="Species Biodiversity", y="Human Pressure")

pB <- pressures2 %>% 
   ggplot(aes(x=percent_GDP, y=percent_MPA)) +
   scale_y_continuous(trans = "log10") +
   scale_x_continuous(trans = 'log10') +
   geom_point() +
   geom_smooth(method='lm',formula=y~x) +
   theme_light() +
   labs(x="% of World GDP", y=" % MPA/EEZ") 

target <- data.frame(x=c(-Inf,-Inf,30, 30), y=c(-Inf,Inf ,Inf, -Inf), t=c('b', 'b', 'b','b'))

pC <- pressures %>% 
   select(country, eez_area, mpa, gdp_tot, biodiversity_points) %>% 
   na.omit() %>%
   mutate(percent_MPA = (mpa/eez_area)*100) %>% 
   ggplot(aes(x=percent_MPA, y=biodiversity_points)) +
   geom_polygon(data=target, mapping=aes(x=x, y=y, group=t), alpha=0.2) +
   geom_point() +
   theme_light() +
   labs(x="% MPA/EEZ", y="Species Biodiversity")

cowplot::plot_grid(pA, pB, pC, labels="AUTO", nrow = 1)

```




## Figure XXX
```{r}
protected_areas <- read_excel("data/BP-10-Dataset.xlsx", sheet = "protected_areas")


protected_areas %>%
        filter(habitat_area > 0) %>% 
        mutate(pp = (protected/habitat_area)*100) %>% 
        mutate(pp_to_EEZ = (protected/eez_area)*100) %>% 
        mutate(relative_to_EEZ = (habitat_area/eez_area)*100) %>% 
        group_by(habitat) %>% 
        mutate(world_hab_habitat_area = sum(habitat_area)) %>% 
        ungroup() %>% 
        mutate(relative_habitat_area_to_world = ((habitat_area/world_hab_habitat_area)*100)) %>% 
        mutate_if(is.numeric, round, digits = 2) %>% 
        select(habitat, country, pp) %>% 
        group_by(habitat) %>% 
        mutate(mean_pp = mean(pp), 
               median_pp = median(pp)) %>% 
        ungroup() %>% 
        mutate(habitat = factor(.$habitat, labels=c("Cold Corals", "Coral Reefs",
                                                          "Estuaries", "Hydrothermal vents", 
                                                          "Kelp", "Mangroves", "Ridges", 
                                                          "Saltmarsh", "Seagrasses", 
                                                          "Seamounts guyots", "Shelf valley canyons", 
                                                          "Trenches"))) %>% 
        ggplot(aes(x=pp))+
        geom_histogram(fill="grey60", col="black")+
        geom_vline(aes(xintercept=mean_pp), col="blue", size=1)+
        geom_vline(aes(xintercept=median_pp), col="red", size=1)+
        geom_vline(aes(xintercept=30), col="black", linetype="dashed")+
        labs(x= "% protected", y="Count")+
        facet_wrap(~habitat)+
        theme_tufte()+
        theme(text = element_text(size = 15),
              panel.border = element_rect(fill = NA, colour = "grey60"))

```


# References 


Sala, E., Mayorga, J., Costello, C., Kroodsma, D., Palomares, M. L. D., Pauly, D., […] Zeller, D. 
(2018). The economics of fishing the high seas. Science Advances, 4(6), eaat2504. https://doi.org/10.1126/sciadv.aat2504

## Habitat data references from the table

- Estuaries: Alder J (2003). Putting the coast in the “Sea Around Us”. The Sea Around Us Newsletter 15: 1-2. URL: http://seaaroundus.org/newsletter/Issue15.pdf; http://data.unep-wcmc.org/datasets/23 (version 2.0) 

- Mangroves: Giri, C., E. Ochieng, L. L. Tieszen, Z. Zhu, A. Singh, T. Loveland, J. Masek, and N. Duke. (2011). “Status and Distribution of Mangrove Forests of the World Using Earth Observation Satellite Data: Status and Distributions of Global Mangroves.” Global Ecology and Biogeography 20 (1): 154–59. https://doi.org/10.1111/j.1466-8238.2010.00584.x.

- Saltmarsh: Mcowen C, Weatherdon LV, Bochove J, Sullivan E, Blyth S, Zockler C, Stanwell-Smith D, Kingston N, Martin CS, Spalding M, Fletcher S (2017). A global map of saltmarshes. Biodiversity Data Journal 5: e11764. Paper DOI: https://doi.org/10.3897/BDJ.5.e11764; Data URL: http://data.unep-wcmc.org/datasets/43 (v.6)

- Seagrasses:UNEP-WCMC, Short FT (2018). Global distribution of seagrasses (version 6.0). Sixth update to the data layer used in Green and Short (2003). Cambridge (UK): UN Environment World Conservation Monitoring Centre. URL: http://data.unep-wcmc.org/datasets/7

- Coral Reefs: UNEP-WCMC, WorldFish Centre, WRI, TNC (2018). Global distribution of warm-water coral reefs, compiled from multiple sources including the Millennium Coral Reef Mapping Project. Version 4.0. Includes contributions from IMaRS-USF and IRD (2005), IMaRS-USF (2005) and Spalding et al. (2001). Cambridge (UK): UN Environment World Conservation Monitoring Centre. URL: http://data.unep-wcmc.org/datasets/1

- Kelp: Jorge Assis (submitted for publication) 

- Coldcorals: Freiwald A, Rogers A, Hall-Spencer J, Guinotte JM, Davies AJ, Yesson C, Martin CS, Weatherdon LV (2017). Global distribution of cold-water corals (version 5.0). Fifth update to the dataset in Freiwald et al. (2004) by UNEP-WCMC, in collaboration with Andre Freiwald and John Guinotte. Cambridge (UK): UN Environment World Conservation Monitoring Centre. URL: http://data.unep-wcmc.org/datasets/3

- Sills-Rift Valleys: Harris, P. T., Macmillan-Lawler, M., Rupp, J., & Baker, E. K. (2014). Geomorphology of the oceans. Marine Geology, 352, 4–24. https://doi.org/10.1016/j.margeo.2014.01.011
Hydrothermal vents: Beaulieu, S.E., Szafranski, K. (2018) InterRidge Global Database of Active Submarine Hydrothermal Vent Fields, Version 3.4. World Wide Web electronic publication available from http://vents-data.interridge.org Accessed 2019-02-20.

# Contacts

- Fabio Favoretto: favoretto.fabio@gmail.com

- Joy Kumagai: jkumagai96@gmail.com

- Octavio Aburto: maburto@ucsd.edu